Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs

### 2014-10-17 papers/termination/termination-esop/paper.tex
title change
abstract + fixes in introduction + kalashnikov citation

\title{Termination and Non-Termination Proofs for Bit-Vector Programs}

@@

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
ranking function for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of functions is
insufficient for proving termination of many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

We present a reduction from program \emph{termination} to 
second-order satisfiability. 
%program
%\emph{synthesis} via a fragment of second-order logic.  
We give formulations for termination and non-termination in a fragment of second-order logic 
with restricted quantification.

existential second-order logic over finite domains is decidable \cite{kalashnikov}.

% decidable for finite-state programs

This reduction
allows us to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic.  We use this reduction to build a sound and
complete procedure for the termination analysis of finite-state programs with fixed-width
integers and IEEE floating-point arithmetic.
%
\end{abstract}

@@

We present a technique that treats uniformly linear and non-linear programs and
it is not restricted to finding linear ranking functions, but can
also compute lexicographic nonlinear ones. 
Our approach is constraint-based and relies on second-order formulations 
of termination and non-termination.
The obvious issue is that,
due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
To make solving our constraints tractable, we formulate termination and non-termination inside a fragment of second-order logic with restricted quantification, 
for which we have built a solver in \cite{}.
Our method is sound and
complete for bit-vector programs -- for any program, we find
a proof of either its termination or non-termination.

@@

\item  We rephrased the termination and non-termination problems as
second-order $\exists \forall$ satisfaction problems.  This formulation
captures the (non-)termination
properties of all of the loops in the program, including
nested loops.  We can use this to analyse all the loops at once,
or one at a time.  Our treatment handles termination and non-termination uniformly:
both properties are captured in the same second-order formula.

@@

\item We implemented our technique and tried it on a selection of programs
handling both bit-vectors and floats. In our implementation we made use of 
a solver for second-order logic with restricted quantification that we have introduced in \cite{}.

@@

Due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
In \cite{}, we have identified and built a solver for a fragment of second-order logic with restricted quantification, 
which we call second-order SAT (see Definition~\ref{def:2sat}).

>>>>

\title{Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs}

@@

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
ranking function for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of functions is
insufficient for proving termination of many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.
We propose a termination analysis able 
to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic
Our technique is based on a reduction from program \emph{termination} to 
second-order satisfaction. Thus, we provide 
formulations for termination and non-termination in a fragment of second-order logic 
with restricted quantification which was proven decidable over finite domains.
%existential second-order logic over finite domains is decidable \cite{kalashnikov}. 
The resulted technique is a sound and complete analysis for the termination 
of finite-state programs with fixed-width
integers and IEEE floating-point arithmetic.
%
\end{abstract}

@@

We present a technique that treats uniformly linear and non-linear programs and
it is not restricted to finding linear ranking functions, but can
also compute lexicographic nonlinear ones. 
Our approach is constraint-based and relies on second-order formulations 
of termination and non-termination.
The obvious issue is that,
due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
To make solving our constraints tractable, we formulate termination and non-termination inside a fragment of second-order logic with restricted quantification, 
for which we have built a solver in \cite{kalashnikov}.
Our method is sound and
complete for bit-vector programs -- for any program, we find
a proof of either its termination or non-termination.

@@

\item  We rephrased the termination and non-termination problems as
second-order %$\exists \forall$ 
satisfaction problems.  This formulation
captures the (non-)termination
properties of all of the loops in the program, including
nested loops.  We can use this to analyse all the loops at once,
or one at a time.  Our treatment handles termination and non-termination uniformly:
both properties are captured in the same second-order formula.

@@

\item We implemented our technique and tried it on a selection of programs
handling both bit-vectors and floats. In our implementation we made use of 
a solver for a fragment of second-order logic with restricted quantification that is decidable 
over finite domains \cite{kalashnikov}.

@@

Due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
In \cite{kalashnikov}, we have identified and built a solver for a fragment of second-order logic with restricted quantification, 
which we call second-order SAT (see Definition~\ref{def:2sat}).
